---
title: "Choice_analysis"
author: "Martine Lind Jensen"
date: "2025-04-21"
output: html_document
---

```{r loading packages}
pacman::p_load(tidyverse, ordinal, lmerTest, smplot2, simr)
```

```{r loading data}
data <- read_csv("data_for_analysis/why_play_is_fun_data2.csv") %>% dplyr::select(-c("...1"))
```

#Modeling choice 

Choice is a binary variable of which of the hard and the easy puzzle the participant most wanted to play with. Choice is made as an extra outcome variable to check if the ratings were reliable. 

Choice is the same for trial, therefore we need to change the dataframe to wide format to make sense in modelling choice. 

```{r}
data_choice <- data %>% 
  dplyr::select(-c(Puzzle_no)) %>% 
  mutate(Forced_choice_trial = ifelse(is.na(Forced_choice_trial), "NA", Forced_choice_trial)) %>% 
  drop_na(Time) %>% 
  pivot_wider(names_from = Diff, values_from = c(Rating, Time, Start, End, Time_hms)) %>% #Making long format for Diff being per trial
  group_by(ID) %>% 
  mutate(Choice = as.factor(dplyr::lead(Pref_trial)), #Creating choice as choice after a trial of E and H. 
         First_choice = dplyr::first(Pref_trial, order_by = Trial)) %>%  #Creating first choice as a parameter from the pref_trial column. 
  ungroup() %>% 
  mutate(
    Highest_rating = ifelse(Rating_E > Rating_H, "E", #Creating highest rating column 
                            ifelse(Rating_H > Rating_E, "H", 
                                   ifelse(Rating_H == Rating_E, Forced_choice_trial, NA))), 
    Age_scaled = scale(Age), 
    Time_E_scaled = scale(Time_E), 
    Time_H_scaled = scale(Time_H), 
    Rating_diff = Rating_H - Rating_E) %>%
  drop_na(Choice)



data_choice <- data_choice %>% 
  filter(!(ID %in% c("hk","hr", "uo", "1t"))) %>% #removing hk, hr and uo as they only have 1 trial and 1t since they didnt do the forced choice
  filter(!(ID == "8l" & Trial == 3)) #removing 8l trial 3, as the participant didnt finish trial 3

```

Checking to see the code worked as intended. 
```{r}
ip <- data_choice %>% filter(ID== "ip")


data_choice %>% filter(Trial == 2) %>% view("Pref_trial")
```

I am modeling choice in three ways. 
(1) Choice as internal validity, relation with rating. 
(2) Choice as a possibility for decision making model, i.e. checking parameter effect on choice outcome. 
(3) Choice as an outcome, i.e. did participant tend to choose hard?

##Choice and rating 

```{r}
data_choice <- data_choice %>% mutate(Rating_diff = Rating_H - Rating_E)

#Modelling with difference of ratings but ordered. 
rating_ord <- glmer(factor(Choice) ~ ordered(Rating_diff) + (1|ID), family = binomial(link = "logit"), data = data_choice, control = glmerControl(optimizer = "bobyqa"))

  #Convergence issues, too complex for the small dataset 

#Modelling as numeric, wrongful model to compare with. 
rating_num <- glmer(factor(Choice) ~ Rating_diff + (1|ID), family = binomial(link = "logit"), data = data_choice)


#Modeling with highest rating, most proper way when the predictor is ordered
highest_rating <- glmer(Choice ~ Highest_rating + (1|ID), family = binomial(link = "logit"), data = data_choice)

anova(rating_ord, highest_rating, rating_num)

summary(highest_rating)
```

Exploration with full dataset. 
```{r}
#Full datase models for testing
highest_rating_full <- glmer(factor(Pref_trial) ~ Highest_rating + (1|ID), family = binomial(link = "logit"), data = data_choice)

rating_full <- glmer(factor(Pref_trial) ~ I(Rating_H - Rating_E) + + (1|ID), family = binomial(link = "logit"), data = data_choice)

anova(highest_rating_full, rating_full)
```

Prediction with the model of highest rating
```{r}
prediction1 <- data.frame(ggpredict(highest_rating, terms = c("Highest_rating")))

ggplot(prediction1, aes(x = x, y = predicted)) + 
  geom_point(aes(color = x), size = 2, position =position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = x), position = position_dodge(width = 0.5), width = 0.3) + 
  theme_minimal()+ 
  scale_color_manual(values=c("#00A08A", "#FF0000")) + 
  ggtitle("Predicted Choices by Highest Rating") + 
  ylab("Predicted Choice of Hard") + 
  xlab("Highest Rating") + 
  theme(legend.position = "none")
```

##Other parameters that predicts choice (decision making model exploration)
```{r}
#Does time spend on the puzzle type predict choice? 
time <- glmer(factor(Choice) ~ I(Time_H_scaled - Time_E_scaled) + (1|ID), family = binomial(link = "logit"), data = data_choice)

#Does base smiley predict choice? 
base_smiley <- glmer(factor(Choice) ~ ordered(Base_smiley) + (1|ID), family = binomial(link = "logit"), data = data_choice)

#Is the first choice a good predictor of choice 
first_choice <- glmer(factor(Choice) ~ factor(First_choice) + (1|ID), family = binomial(link = "logit"), data = data_choice)

#Does the previous choice predict the following choice? 
prev_choice <- glm(factor(Choice) ~ Pref_trial, family = binomial(), data = data_choice)

#Any of the descriptive statistics that predicts choice? 
descr <- glmer(factor(Choice) ~ Age_scaled + Gender + (1|ID), family = binomial(), data = data_choice)

full <- glmer(factor(Choice) ~ I(Time_H_scaled - Time_E_scaled) + ordered(Base_smiley) + factor(First_choice) + factor(Pref_trial) + Age_scaled + Gender + (1|ID), family = binomial(), data = data_choice, control = glmerControl(optimizer="bobyqa"))

vif(full) #Eventhough there are some significant parameters in the full model, it is highly correlated, so not a good idea to use it. 
```

Predictions and plots for previous choice model
```{r}
prediction2 <- data.frame(ggpredict(prev_choice, terms = c("Pref_trial")))

ggplot(prediction2, aes(x = x, y = predicted)) + 
  geom_point(aes(color = x), size = 2, position =position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = x), position = position_dodge(width = 0.5), width = 0.3) + 
  theme_minimal()+ 
  scale_color_manual(values=c("#5BBCD6", "#F98400")) + 
  ggtitle("Predicted Choices by Previous Choices") + 
  ylab("Predicted Choice of Hard") + 
  xlab("Previous Choice") + 
  theme(legend.position = "none") + 
  ylim(0,1)


```

##Difference of choice
```{r}
#Are there a significant difference in choice? 
null <- glmer(factor(Choice) ~ 1 + (1|ID), family = binomial(), data = data_choice)

summary(null)

plogis(fixef(null)) #69,72446 % chose hard 

null_full <- glmer(factor(Pref_trial) ~ 1 + (1|ID), family = binomial(), data = data) #with all participants

summary(null_full)

plogis(fixef(null_full)) #0.5456262% chose H
```


##Did the participants change their preference choice after 1. round? 
```{r}
#Filtering only trial 1

df_switch <- data_choice %>% filter(Trial == 1) %>% mutate(Switch_after_first = ifelse(First_choice == Choice, 0, 1))

switch <- glm(Switch_after_first ~ 1, data = df_switch, family = binomial())

switch_easy <- glm(Switch_after_first ~ 1, data = filter(df_switch, First_choice == "E"), family = binomial())

summary(switch_easy)
#It is significantly different from 50% but - means that it is because the 0 value, no switch is 78% 

plogis(-1.204)
plogis(-0.5878)
```


```{r squarred}
MuMIn::r.squaredGLMM(null)
#marginal = fixed effects only, condition fixed + random
```



## Plots
```{r}
data_choice %>% ggplot(aes(x= factor(Pref_trial), Highest_rating), fill = Highest_rating) + 
  geom_point() + 
  facet_wrap(.~ID)

data_choice %>% drop_na() %>% ggplot() + 
  geom_jitter(aes(Rating_E, Choice), color = "red") + 
  geom_jitter(aes(Rating_H, Choice), color = "blue") + 
  xlab("Rating")


data_choice %>% ggplot() + 
  geom_jitter(aes(Rating_E, Pref_trial), color = "red") + 
  geom_jitter(aes(Rating_H, Pref_trial), color = "blue") + 
  geom_violin(aes(Rating_E, Pref_trial), fill = "red3", alpha =0.5) + 
  geom_violin(aes(Rating_H, Pref_trial), fill = "lightblue", alpha = 0.5)



data_choice %>% drop_na() %>% ggplot() + 
  geom_jitter(aes(Rating_E, Choice), color = "red") + 
  geom_jitter(aes(Rating_H, Choice), color = "blue") + 
  sm_violin(mapping = aes(Rating_E, Choice), color = "red3", alpha =0.5, point.params = list(size = 1.5)) + 
  sm_violin(mapping = aes(Rating_H, Choice), fill = "lightblue", alpha = 0.5, point.params = list(size = 1.5))
```

```{r}
ggplot(data_choice, aes(x = Rating_E, y = Choice)) +
  geom_jitter(height = 0.05, width = 0.1, alpha = 0.3) +
  stat_smooth(method = "glm", method.args = list(family = "binomial"),
              se = TRUE, color = "steelblue") +
  #scale_y_continuous(breaks = c(0, 1), labels = c("Easy", "Hard")) +
  labs(
    title = "Logistic Regression: Choice vs. Enjoyment Rating",
    x = "Enjoyment Rating (1–5)",
    y = "Probability of Choosing 'Hard'"
  ) +
  theme_minimal()
```

```{r}
plot(ggpredict(highest_rating, terms = c("Highest_rating")))

ggpredict(rating_num) %>% plot()

ggpredict(rating_ord) %>% plot()

ggpredict(prev_choice) %>% plot()

```


##Power analysis since it seems like there might be an effect of highest rated anyways 

```{r}
# Extend to 100 participants
extended_model <- extend(highest_rating, within = "Trial+ID", n = 50)
extended_model
# Simulate power
power_result <- powerSim(highest_rating, nsim = 100)  # 100 simulations
power_result


powers <- powerCurve(extended_model, within = "Trial+ID", nsim = 10)

powers$warnings

plot(powers)
```

```{r}
summary_df <- data_choice %>%
  group_by(Rating_E) %>%
  summarise(
    prob_hard = mean(Choice),
    n = n(),
    se = sqrt(prob_hard * (1 - prob_hard) / n),
    ci_low = prob_hard - 1.96 * se,
    ci_high = prob_hard + 1.96 * se
  )

# Plot
ggplot(summary_df, aes(x = rating, y = prob_hard)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2, color = "steelblue") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Probability of Choosing 'Hard' by Enjoyment Rating",
    x = "Enjoyment Rating (1–5)",
    y = "Probability of Choosing 'Hard'"
  ) +
  theme_minimal()
```


```{r}
# Get fixed effect (intercept) and standard error
est <- fixef(null)[1]
se <- sqrt(vcov(null)[1, 1])

# Convert log-odds to probability
library(boot)
p <- inv.logit(est)
ci <- inv.logit(c(est - 1.96 * se, est + 1.96 * se))

```

```{r}
library(ggplot2)

df_plot <- data.frame(
  label = "Overall",
  probability = p,
  lower = ci[1],
  upper = ci[2]
)

ggplot(df_plot, aes(x = label, y = probability)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, color = "steelblue") +
  ylim(0, 1) +
  ylab("Predicted Probability of Choosing Hard") +
  xlab("") +
  ggtitle("Overall Preference for 'Hard' Choice") +
  theme_minimal()

```

```{r}
ranef_vals <- ranef(null)$ID[[1]] + fixef(null)[1]
subject_ids <- rownames(ranef(null)$ID)

subject_probs <- inv.logit(ranef_vals)

df_subjects <- data.frame(
  participant = subject_ids,
  probability = subject_probs
)


ggplot(df_subjects, aes(x = reorder(participant, probability), y = probability)) +
  geom_point(color = "gray40") +
  geom_hline(yintercept = boot::inv.logit(fixef(null)[1]), color = "steelblue", linetype = "dashed") +
  coord_flip() +
  ylim(0,1) +
  ylab("Predicted Probability (Hard)") +
  xlab("Participant") +
  ggtitle("Participant-Level Predicted Probabilities") +
  theme_minimal()

```

```{r}
# Extract fixed effect (intercept) and standard error
intercept <- fixef(null)[1]
se <- sqrt(vcov(null)[1,1])
ci <- intercept + c(-1.96, 1.96) * se

# Create a sequence of log-odds around the intercept
log_odds <- seq(intercept - 5, intercept + 5, length.out = 200)
probabilities <- inv.logit(log_odds)

# Prepare data for plotting
df_plot <- data.frame(
  log_odds = log_odds,
  probability = probabilities
)

# Plot logistic curve with estimated intercept and CI
ggplot(df_plot, aes(x = log_odds, y = probability)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_vline(xintercept = intercept, linetype = "dashed", color = "black") +
  geom_ribbon(
    data = subset(df_plot, log_odds >= ci[1] & log_odds <= ci[2]),
    aes(ymin = 0, ymax = probability),
    fill = "steelblue", alpha = 0.2
  ) +
  labs(
    title = "Logistic Curve from Intercept-Only Model",
    x = "Log-Odds (Linear Predictor)",
    y = "Predicted Probability of Choosing 'Hard'"
  ) +
  theme_minimal()
```

```{r}
 #Compute model estimate and CI on probability scale
prob <- boot::inv.logit(intercept)
ci_prob <- boot::inv.logit(ci)

# Add jittered choice values to raw data
data_choice$jittered_choice <- jitter(data_choice$Choice, amount = 0.05)

# Plot 2: Binary outcome vs predicted probability
ggplot(data_choice, aes(x = prob, y = Choice)) +
  geom_jitter(width = 0.01, height = 0.05, alpha = 0.3, color = "gray40") +
  geom_point(aes(y = prob), size = 4, color = "steelblue") +
  geom_errorbarh(aes(xmin = ci_prob[1], xmax = ci_prob[2], y = prob),
                 height = 0.05, color = "steelblue") +
  #scale_y_continuous(breaks = c(0, 1), labels = c("Easy", "Hard")) +
  labs(
    title = "Estimated Probability of Choosing 'Hard'",
    x = "Predicted Probability",
    y = "Observed Binary Choice"
  ) +
  theme_minimal()
```

